Contrastive Optimization -- Preliminaries

### 2024-09-27 paper/main.tex
added ablated cross entropy formulation to paper, normalized fg/bg split influence plots to full heatmap

\begin{theorem}
	Individual class probabilites $\sigma_c(\bm{s})$ depend on the difference(s) between class logits.
\end{theorem}
\begin{proof} Individual class probabilities for logit vector $\bm{s}$ are defined as:
	\begin{gather}
		\sigma_c(\bm{s}) = \frac{e^{s_c}}{\sum_i e^{s_i}}
	\end{gather}
	We define our logit vector in terms of the elementwise difference to a target class $c$:
	\begin{gather}
		d = [s_i - s_c]_i \\
		\bm{s} = [s_c + d_i]_i
	\end{gather}
	Based on this definition, class probabilities can now be computed as:
	\begin{align}
		\sigma_c(\bm{s}) &= \frac{e^{s_c}}{\sum_i e^{s_c+d_i}} \\
				 &= \frac{e^{s_c}}{e^{s_c} \sum_i e^{d_i}} = \frac{1}{\sum_i e^{d_i}}
	\end{align}
	This recontextualizes softmax as a direct function of the differences of class logits.
\end{proof}
\begin{corollary} Since class probabilities depend on logit differences, the class probabilities are identitical for any two pairs of logit vectors with identical logit differences.
	\begin{gather}
		\frac{e^{s_c}}{\sum_i e^{s_i}} = \frac{e^{s_c+r}}{\sum_i e^{s_i+r}} \forall r \in \mathbb{R}
	\end{gather}
\end{corollary}
\begin{proof} Recontextualizing $\bm{s}$ per the previous proposition, we get:
	\begin{align}
		\frac{e^{s_c}}{\sum_i e^{s_c + d_i}} &= \frac{e^{s_c+r}}{\sum_i e^{s_c+d_i+r}} \\
		\frac{e^{s_c}}{e^{s_c}\sum_i e^{d_i}} &= \frac{e^{s_c+r}}{e^{s_c + r} \sum_i e^{d_i}} \\
		\frac{1}{\sum_i e^{d_i}} &= \frac{1}{\sum_i e^{d_i}}
	\end{align}
	Therefore, the absolute scale of a logit does not influence the final class probability.
\end{proof}

@@

Segmentations maps describe the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
\begin{gather}
	\{(x, (h, y))\}^N \in \mathcal{D} \text{ where } h_{i,j} := \begin{cases}
		1 & \text{pixel contains subject} \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{background}}_{c,i,j} := \begin{cases}
		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & h_{i,j} = 0 \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{foreground}}_{c,i,j} := \begin{cases}
		0 & h_{i,j} = 0 \\
		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & \text{otherwise}
	\end{cases} \\
	D_{KL}(P || Q) := \frac{1}{|c|-1}\sum^{|c|-1}_c P \cdot \log \frac{P'}{Q_c} \text{ where } 
	P'_{i,j} := \begin{cases}
		1 & h_{i,j} \neq 0 \\
		P_{i,j} & \text{otherwise}
	\end{cases} \label{d_kl} \\
	\begin{split}
		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \mathcal{\bm{A}}^{\text{contrastive}})) \\
		&+ \frac{\lambda_4}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 0)} ||\mathcal{\bm{A}}^{\text{background}}||^2_F \\
		&- \frac{\lambda_5}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 1)} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
	\end{split}
\end{gather}

The cost function has a threefold purpose:
\begin{enumerate}
	\item Minimize the divergence of the contrastive activation map to the segmentation map.
	\item Minimize the absolute value of the contrastive map where the segmentation map is the background.
	\item Maximize the value of the contrastive map where the segmentation map is the foreground.
\end{enumerate}

@@

@@

\section{Results}

@@

\vspace{-2em}
\begin{figure}[H]
	\centering
	\includegraphics[width=.4\textwidth]{img/accuracy.png}
	\hspace{4em}
	\includegraphics[width=.4\textwidth]{img/loss.png}
	\caption{Plots comparing benchmark cross-entropy loss, training and validation accuracy. The web version of this comparison can be found \href{https://dagshub.com/jinensetpal/contrastive-optimization/experiments\#/compare?experiments=[\%22m_f4f7dea52fa74739be788e5f47da1195\%22,\%22m_1301bf1ba9bf4db48e09e7497c8cd214\%22]}{here}. The blue line represents cross-entropy loss, while the orange line represents the contrastive approach.}
\end{figure}

>>>>

\begin{theorem}\label{classprobdiff}
	Individual class probabilites $\sigma_c(\bm{s})$ depend on the difference(s) between class logits.
\end{theorem}
\begin{proof} Individual class probabilities for logit vector $\bm{s}$ are defined as:
	\begin{gather}
		\sigma_c(\bm{s}) = \frac{e^{s_c}}{\sum_i e^{s_i}}
	\end{gather}
	We define our logit vector in terms of the elementwise difference to a target class $c$:
	\begin{gather}
		d = [s_i - s_c]_i \\
		\bm{s} = [s_c + d_i]_i
	\end{gather}
	Based on this definition, class probabilities can now be computed as:
	\begin{align}
		\sigma_c(\bm{s}) &= \frac{e^{s_c}}{\sum_i e^{s_c+d_i}} \\
				 &= \frac{e^{s_c}}{e^{s_c} \sum_i e^{d_i}} = \frac{1}{\sum_i e^{d_i}}
	\end{align}
	This recontextualizes softmax as a direct function of the differences of class logits.
\end{proof}
\begin{corollary} Since class probabilities depend on logit differences, the class probabilities are identitical for any two pairs of logit vectors with identical logit differences.
	\begin{gather}
		\frac{e^{s_c}}{\sum_i e^{s_i}} = \frac{e^{s_c+r}}{\sum_i e^{s_i+r}} \forall r \in \mathbb{R}
	\end{gather}
\end{corollary}
\begin{proof} Recontextualizing $\bm{s}$ per the previous proposition, we get:
	\begin{align}
		\frac{e^{s_c}}{\sum_i e^{s_c + d_i}} &= \frac{e^{s_c+r}}{\sum_i e^{s_c+d_i+r}} \\
		\frac{e^{s_c}}{e^{s_c}\sum_i e^{d_i}} &= \frac{e^{s_c+r}}{e^{s_c + r} \sum_i e^{d_i}} \\
		\frac{1}{\sum_i e^{d_i}} &= \frac{1}{\sum_i e^{d_i}}
	\end{align}
	Therefore, the absolute scale of a logit does not influence the final class probability.
\end{proof}

@@

\subsubsection{Divergence Formulation}
Segmentations maps describe the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
\begin{gather}
	\{(x, (h, y))\}^N \in \mathcal{D} \text{ where } h_{i,j} := \begin{cases}
		1 & \text{pixel contains subject} \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{background}}_{c,i,j} := \begin{cases}
		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & h_{i,j} = 0 \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{foreground}}_{c,i,j} := \begin{cases}
		0 & h_{i,j} = 0 \\
		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & \text{otherwise}
	\end{cases} \\
	D_{KL}(P || Q) := \frac{1}{|c|-1}\sum^{|c|-1}_c P \cdot \log \frac{P'}{Q_c} \text{ where } 
	P'_{i,j} := \begin{cases}
		1 & h_{i,j} \neq 0 \\
		P_{i,j} & \text{otherwise}
	\end{cases} \label{d_kl} \\
	\begin{split}
		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \underbrace{\lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \mathcal{\bm{A}}^{\text{contrastive}}))}_{\text{minimize divergence from target shape}} \\
		&+ \underbrace{\frac{\lambda_4}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 0)} ||\mathcal{\bm{A}}^{\text{background}}||^2_F}_{\text{minimize background influence}} \\
		&- \underbrace{\frac{\lambda_5}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 1)} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}}_{\text{maximize foreground influence}}
	\end{split}
\end{gather}

@@

\subsubsection{Ablated Cross-Entropy}

A more mathematically rigorous approach involves performing an ablation over class logits to penalize background influence.

\begin{theorem} We can integrate background invariance to the definition of cross-entropy using the following formulation:
	\begin{gather}
		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
	\end{gather}
\end{theorem}
\begin{proof} Using the definition of HiResCAMs, ContrastiveCAMs and the elementwise difference setup from \ref{classprobdiff}, we have:
	\begin{align}
		d &= [s_i - s_c]_i \\
		&= \left[\sum^{D_1,D_2}_{i,j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{c,j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right] \\
		&= \left[ -\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right]_i
	\end{align}
	Using our result from \ref{classprobdiff}, we have:
	\begin{gather}
		\sigma_c(\bm{s}) = \frac{1}{\sum_i e^{d_i}} = \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})}
	\end{gather}
	For a target class $c$, cross-entropy loss is defined as:
	\begin{gather}
		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})} \right)
	\end{gather}
	We can break down each contrastive map to foreground and background components:
	\begin{gather}
		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} - \sum (1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}\right)} \right)
	\end{gather}
	The current loss formulation motivates improving contrast using either the foreground or the background. We penalize contrast on the background, motivating learning invariance to noise in the form of the background.
	\begin{gather}
		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
	\end{gather}
	This gives us our ablated cross-entropy formulation.
\end{proof}

@@

\section{Results [OUTDATED, IGNORE FOR NOW]}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=.4\textwidth]{img/accuracy.png}
	\hspace{4em}
	\includegraphics[width=.4\textwidth]{img/loss.png}
	\caption{Plots comparing benchmark cross-entropy loss, training and validation accuracy. The web version of this comparison can be found \href{https://dagshub.com/jinensetpal/contrastive-optimization/experiments\#/compare?experiments=[\%22m_f4f7dea52fa74739be788e5f47da1195\%22,\%22m_1301bf1ba9bf4db48e09e7497c8cd214\%22]}{here}. The blue line represents cross-entropy loss, while the orange line represents the contrastive approach.}
\end{figure}
